---
title: "copulaStan: Fitting Bivariate Gaussian and Clayton Copulas"
author: "Benedikt Lugauer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{copulaStan: Fitting Bivariate Gaussian and Clayton Copulas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In this vignette, we demonstrate how to fit bivariate Gaussian copula models using the `copulaStan` package. We provide examples for fitting Gaussian copula models with different types of marginals, including normal, and lognormal distributions.

```{r}
library(copulaStan)
library(copula)
```

## Example: Gaussian Copula with Normal and Lognormal Marginals

```{r, eval=FALSE}
library(copulaStan)
library(copula)

seed <- 123
set.seed(seed)
true_rho <- 0.5
n <- 2000  

margins <- c("norm", "lnorm")
params <- list(list(mean = 0.8, sd = 2), list(meanlog = 0, sdlog = 0.8))
cop <- normalCopula(param = true_rho, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_gaussian_copula(data, marginals = c("normal", "lognormal"), 
                           iter = 4000, chains = 4, warmup = 2000,
                           seed = seed, cores = 4)
print(fit$fit)
```

## Derivation of Gaussian Copula Log Likelihood

Here is a derivation of the Gaussian copula likelihood implemented in this package.

**Transformation to Uniform Marginals**:
\[
U_1 = F_1(X_1), \quad U_2 = F_2(X_2)
\]
where \(F_1\) and \(F_2\) are the CDFs of the marginals.

**Transformation to Standard Normal**:
\[
Z_1 = \Phi^{-1}(U_1), \quad Z_2 = \Phi^{-1}(U_2)
\]
where \(\Phi^{-1}\) is the inverse CDF of the standard normal distribution.

**Bivariate Normal Density**:
\[
\phi_{\rho}(z_1, z_2) = \frac{1}{2\pi\sqrt{1-\rho^2}} \exp\left( -\frac{1}{2(1-\rho^2)}(z_1^2 + z_2^2 - 2\rho z_1 z_2) \right)
\]

**Standard Normal Marginal Density**:
\[
\phi(z) = \frac{1}{\sqrt{2\pi}} \exp\left( -\frac{z^2}{2} \right)
\]

**Copula Density**:
\[
c(u_1, u_2; \rho) = \frac{\phi_{\rho}(z_1, z_2)}{\phi(z_1) \phi(z_2)}
\]
Substituting the expressions for \(\phi_{\rho}\) and \(\phi\):
\[
c(u_1, u_2; \rho) = \frac{\frac{1}{2\pi\sqrt{1-\rho^2}} \exp\left( -\frac{1}{2(1-\rho^2)}(z_1^2 + z_2^2 - 2\rho z_1 z_2) \right)}{\left( \frac{1}{\sqrt{2\pi}} \exp\left( -\frac{z_1^2}{2} \right) \right) \left( \frac{1}{\sqrt{2\pi}} \exp\left( -\frac{z_2^2}{2} \right) \right)}
\]
Simplifying, we get:
\[
c(u_1, u_2; \rho) = \frac{1}{\sqrt{1-\rho^2}} \exp\left( -\frac{1}{2(1-\rho^2)}(z_1^2 + z_2^2 - 2\rho z_1 z_2) + \frac{z_1^2}{2} + \frac{z_2^2}{2} \right)
\]

**Log Likelihood**:
\[
\log c(u_1, u_2; \rho) = -\frac{1}{2} \log(1-\rho^2) + \frac{\rho z_1 z_2 - \rho^2 (z_1^2 + z_2^2)}{2(1-\rho^2)}
\]

