---
title: "copulaStan: Fitting Bivariate Copula Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{copulaStan: Fitting Bivariate Copula Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Stan models require CmdStan; don't evaluate during R CMD check
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This vignette demonstrates how to fit bivariate copula models using the
`copulaStan` package. The package supports Gaussian, Clayton, and Joe copulas
with normal, lognormal, exponential, and beta marginal distributions.

```{r}
library(copulaStan)
library(copula)
```

## Example: Gaussian Copula

### Normal + Lognormal Marginals

```{r}
seed <- 2024
set.seed(seed)
true_rho <- 0.5
n <- 2000

margins <- c("norm", "lnorm")
params <- list(list(mean = 0.8, sd = 2), list(meanlog = 0, sdlog = 0.8))
cop <- normalCopula(param = true_rho, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(data,
  copula = "gaussian", marginals = c("normal", "lognormal"),
  seed = seed
)
fit
summary(fit)
coef(fit)
```

### Normal + Exponential Marginals

```{r}
seed <- 2024
set.seed(seed)
true_rho <- 0.5
n <- 2000

margins <- c("norm", "exp")
params <- list(list(mean = 0.8, sd = 2), list(rate = 1))
cop <- normalCopula(param = true_rho, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(data,
  copula = "gaussian", marginals = c("normal", "exponential"),
  seed = seed
)
fit
```

### Beta + Beta Marginals

```{r}
seed <- 2024
set.seed(seed)
true_rho <- 0.5
n <- 2000

margins <- c("beta", "beta")
params <- list(list(shape1 = 2, shape2 = 5), list(shape1 = 3, shape2 = 4))
cop <- normalCopula(param = true_rho, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(data,
  copula = "gaussian", marginals = c("beta", "beta"),
  seed = seed
)
fit
```

## Example: Clayton Copula

### Normal + Lognormal Marginals

```{r}
seed <- 2024
set.seed(seed)
true_theta <- 2.0
n <- 2000

margins <- c("norm", "lnorm")
params <- list(list(mean = 0.8, sd = 2), list(meanlog = 0, sdlog = 0.8))
cop <- claytonCopula(param = true_theta, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(data,
  copula = "clayton", marginals = c("normal", "lognormal"),
  seed = seed
)
fit
```

## Example: Joe Copula

### Normal + Lognormal Marginals

```{r}
seed <- 2024
set.seed(seed)
true_theta <- 2.0
n <- 2000

margins <- c("norm", "lnorm")
params <- list(list(mean = 0.8, sd = 2), list(meanlog = 0, sdlog = 0.8))
cop <- joeCopula(param = true_theta, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(data,
  copula = "joe", marginals = c("normal", "lognormal"),
  seed = seed
)
fit
```


# Mathematical Derivations

## Gaussian Copula Log-Likelihood

**Transformation to Uniform Marginals**:
$$
U_1 = F_1(X_1), \quad U_2 = F_2(X_2)
$$
where $F_1$ and $F_2$ are the CDFs of the marginals.

**Transformation to Standard Normal**:
$$
Z_1 = \Phi^{-1}(U_1), \quad Z_2 = \Phi^{-1}(U_2)
$$

**Copula Density** (ratio of bivariate to product of univariate normals):
$$
c(u_1, u_2; \rho) = \frac{1}{\sqrt{1-\rho^2}} \exp\left( -\frac{1}{2(1-\rho^2)}(z_1^2 + z_2^2 - 2\rho z_1 z_2) + \frac{z_1^2}{2} + \frac{z_2^2}{2} \right)
$$

**Log-Likelihood**:
$$
\log c(u_1, u_2; \rho) = -\frac{1}{2} \log(1-\rho^2) + \frac{\rho z_1 z_2 - \frac{1}{2}\rho^2 (z_1^2 + z_2^2)}{1-\rho^2}
$$

## Clayton Copula Log-Likelihood

The Clayton copula CDF:
$$
C(u, v; \theta) = \left(u^{-\theta} + v^{-\theta} - 1\right)^{-1/\theta}, \quad \theta > 0
$$

The density (second mixed partial derivative):
$$
c(u, v; \theta) = (\theta + 1) \, u^{-\theta - 1} \, v^{-\theta - 1} \left( u^{-\theta} + v^{-\theta} - 1 \right)^{-1/\theta - 2}
$$

Log-likelihood:
$$
\log c(u, v; \theta) = \log(1 + \theta) + (-\theta - 1)(\log u + \log v) - \frac{2\theta + 1}{\theta} \log(u^{-\theta} + v^{-\theta} - 1)
$$

## Joe Copula Log-Likelihood

The Joe copula CDF:
$$
C(u, v; \theta) = 1 - \left[\bar{u}^\theta + \bar{v}^\theta - \bar{u}^\theta \bar{v}^\theta\right]^{1/\theta}, \quad \theta \geq 1
$$
where $\bar{u} = 1-u$, $\bar{v} = 1-v$.

Let $a = \bar{u}^\theta$, $b = \bar{v}^\theta$, $S = a + b - ab$. The density:
$$
c(u, v; \theta) = \theta \, \bar{u}^{\theta-1} \, \bar{v}^{\theta-1} \, S^{1/\theta - 2} \left[ S + \frac{\theta - 1}{\theta}(1 - a)(1 - b) \right]
$$

Log-likelihood:
$$
\log c(u, v; \theta) = \log\theta + (\theta-1)[\log\bar{u} + \log\bar{v}] + \left(\frac{1}{\theta} - 2\right)\log S + \log\left[S + \frac{\theta-1}{\theta}(1-a)(1-b)\right]
$$
