---
title: "Introduction to copulaStan"
author: "Benedikt Lugauer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to copulaStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(copulaStan)
```

# Introduction

This vignette provides an introduction to the `copulaStan` package, which offers functions to fit bivariate Gaussian and Clayton copula models using Stan.

# Installation

To install the development version of `copulaStan` from GitHub, you can use:

```{r}
# install.packages("devtools")
# devtools::install_github("benlug/copulaStan")
```

# Gaussian Copula

The `fit_gaussian_copula` function fits a bivariate Gaussian copula model.

## Example

```{r}
library(copula)
library(copulaStan)

# generate data from a bivariate Gaussian copula
set.seed(123)
n <- 1000
true_rho <- 0.5
cop <- normalCopula(param = true_rho, dim = 2)
U <- rCopula(n, cop)

# fit Bayesian Gaussian copula model
fit <- fit_gaussian_copula(U, 
                           iter = 4000, chains = 4, warmup = 2000,
                           seed = 2024)
fit
```

# Clayton Copula

The `fit_clayton_copula` function fits a bivariate Clayton copula model.

## Example

```{r}
library(copula)
library(copulaStan)

# generate data from a bivariate Clayton copula
set.seed(123)
n <- 1000
true_theta <- 2
cop <- claytonCopula(param = true_theta, dim = 2)
U <- rCopula(n, cop)

# fit the Bayesian Clayton copula model
fit <- fit_clayton_copula(U, 
                          iter = 4000, chains = 4, warmup = 2000, thin = 1, 
                          seed = 2024)

fit
```


